Gammapy: A Python package for gamma-ray astronomy

### 2023-03-08 src/ms-review.tex
implement Matthias comments

\abstract
	{
		Traditionally, TeV-\gammaray astronomy has been conducted
		by experiments employing proprietary data and analysis software.
		However, the next generation of \gammaray instruments,
		such as the the Cherenkov Telescope Array, will be operated as open observatories.
		Alongside the data, they will also make associated software tools available to a wider community.
    This necessity prompted the development of open, high-level astronomy software customised for high-energy astrophysics.
	}
	{
		In this article, we present \gammapy, an open-source Python package for the analysis of astronomical \gammaray data,
		and illustrate the functionalities of its first long-term-support release, version 1.0.
		Built on the modern Python scientific ecosystem, \gammapy provides a uniform platform for reducing and
		modelling data from different \gammaray instruments for many analysis scenarios.
    \gammapy complies with several well-established data conventions in high-energy astrophysics, providing serialised data products that
		are interoperable with other software packages.
	}
	{
		Starting from event lists and instrument response functions,
		\gammapy provides the functionalities for reducing data binned in energy and sky coordinates.
		Several techniques for background estimation are implemented in the package to handle the residual hadronic background.
		After the data are binned, the flux and morphology of one or more \gammaray sources can be estimated
		using Poisson maximum likelihood fitting
		and assuming a variety of spectral, temporal, and spatial models.
		Estimation of flux points, likelihood profiles and light curves is also supported.
	}
	{
		After describing the structure of the package, we show the capabilities of \gammapy
		in multiple traditional and novel \gammaray analysis scenarios using public data
		such as spectral and spectro-morphological modelling and estimations of a spectral energy
		distribution and a light curve.
		Its flexibility and its power are displayed in a final multi-instrument example,
		where datasets from different instruments, at different stages of data reduction,
		are simultaneously fitted with an astrophysical flux model.
	}{}

@@

%% \gammaray astronomy
%% CN: I think we should use the expression Water Cherenkov Detectors and use the
%% abbreviation (WCD) all along the text, later we use "particle samplers"
The \gammaray range of the electromagnetic spectrum provides us insights into the
most energetic processes in the universe such as those accelerating particles in the surroundings of
black holes, and remnants of supernova explosions. As in other
branches of astronomy, \gammarays can be observed by
satellite as well as ground based instruments.
Ground-based instruments use the Earth's atmosphere as a particle detector.
Very-high-energy (VHE) cosmic \gammarays interact in the atmosphere and
create large showers of secondary particles that can be observed from the ground.
Ground-based \gammaray astronomy relies on these extensive air showers to detect the
primary \gammaray photons and infer their incident direction and energy.
VHE \gammaray astronomy covers the energy range from few tens of \si{GeV} up to the \si{PeV}.
There are two main categories of ground-based instruments:

@@

\textit{Water Cherenkov Detectors (WCDs)} detect particles directly from the tail of the
shower when it reaches the ground. These instruments have a very
large FoV and large duty-cycle, but a higher energy threshold and
lower signal-to-noise ratios compared to IACTs~\citep{2015CRPhy..16..610D}.

\begin{figure*}[t]
	\centering
	\includegraphics[width=\textwidth]{figures/big-picture.pdf}
	\caption{
		Core idea and relation of \gammapy to different \gammaray instruments
		and the gamma astro data formats (GADF). The top left shows the
		group of current and future pointing instruments based on the 
		imaging atmospheric Cherenkov technique (IACT). This includes
		instruments such as the Cherenkov Telescope Array (CTA),
		the High Energy Stereoscopic System (H.E.S.S.), the
		Major Atmospheric Gamma Imaging Cherenkov Telescopes (MAGIC),
		and the Very Energetic Radiation Imaging Telescope Array System (VERITAS).
		The lower left shows the group of all-sky instruments such as the
		Fermi Large Area Telescope (Fermi-LAT) and the High Altitude
		Water Cherenkov Observatory (HAWC). The calibrated data of all those
		instruments can be converted and stored into the common GADF data format.
		\gammapy can read data stored in the GADF format.
		The \gammapy package is not a part of any instrument, but instead
		provides a common interface to the data and analysis of all
		these \gammaray instruments. This way users can also easily combine data from
		different instruments and perform a joint analysis.
		\gammapy is built on the scientific Python ecosystem, and the required dependencies
		are shown below the \gammapy logo.
	}
	\label{fig:big_picture}

@@

%TODO: better transition into context
The Cherenkov Telescope Array (CTA) will be the first ground-based \gammaray instrument to be operated as an open observatory.
Its high-level data\footnote{The lowest reduction level of data published by CTAO will be reconstructed event lists and corresponding instrument response functions.} will be shared publicly after
some proprietary period, and the software required to analyze it will be distributed
as well. To allow the re-usability of data from existing instruments and their interoperability,
it is required to use open data formats and open tools that can support the various analysis methods
commonly used in the field.

@@

%% \gammapy: concept and goals
%%% Question: starting date? Is is relevant, what is the real start?
The \gammapy project was started following the model of \astropy, with the objective of building a common
software library for very high-energy \gammaray data analysis \citep{gammapy_2015}. 
The core of the idea is illustrated in Figure~\ref{fig:big_picture}. Various \gammaray instruments
export their data to the standardised common \gadf data format. This data can then be combined and
analysed using a common software library.
This means that the \gammapy package is not specific to any instrument,
but an independent community-developed software project.
The \gammapy package is built on the scientific Python ecosystem: it uses \numpy~\citep{numpy} for n-dimensional data
structures, \scipy~\citep{2020SciPy-NMeth} for numerical algorithms, \astropy~\citep{astropy} for
astronomy-specific functionality, \iminuit~\citep{iminuit} for numerical minimisation
and \matplotlib~\citep{matplotlib} for visualization.

%% CN: what about mentioning the joint Crab? or is it mentioned later?
With the public availability of the \gadf format sepcifications and the
\gammapy package, some experiments started to make limited subsets of
their \gammaray data publicly available for testing and validating
\gammapy. For example, the \hess collaboration released a limited test
dataset (about 50 hours of observations taken between 2004 and 2008)
based on the \gadf DL3 format \citep{HESS_DR1}. This data release served
as a basis for validation of open analysis tools, including \gammapy 
\cite[see e.g.][]{Mohrmann2019}. The \hawc collaboration also released
a limited test dataset of the Crab Nebula, which was used to validate
the \gammapy package in \cite{Olivera2022}.

@@

\section{Gamma-ray Data Analysis}
\begin{figure*}[t]
	\centering
	\includegraphics[width=1.\textwidth]{figures/data_flow.pdf}
	\caption{
		\gammapy sub-package structure and data analysis workflow. The top row
        defines the different levels of data reduction, from lists of \gammaray-like
        events on the left (DL3), to high-level scientific products products
        (DL5) on the right. The direction of the data flow is illustrated with the
        gray arrows. The gray folder icons represent the different sub-packages
        in \gammapy and their names. Below each icon there is a list of the most
        important objects defined in the sub-package.
    }
	\label{fig:data_flow}
\end{figure*}
%
\label{sec:gammaray-data-analysis}
% This might not be the bast place compared to introduction
The data analysis process in \gammaray astronomy is usually split into two parts.
The first one deals with the data processing from camera measurement, calibration, event
reconstruction and selection to yield a list of reconstructed \gammaray event candidates.
This part of the data reduction sequence, sometimes referred to as low-level analysis,
is usually very specific to a given observation technique and even to a given instrument.

The other sequence, referred to as high-level analysis, deals with the extraction of physical
quantities related to \gammaray sources and the production of high-level products such as spectra,
light curves and catalogs. The methods applied here are more generic and are broadly
shared across the field. The similarity in the high-level analysis would also allow
for combining data from multiple instruments, but could not be fully exploited, due to
a lack of common data formats and software tools.

@@

where:
\begin{itemize}
\setlength\itemsep{1em}
\item $A_{\rm eff}(p_{\rm true}, E_{\rm true})$ is the effective collection area of the detector. It is the product
  of the detector collection area times its detection efficiency at true energy $E_{\rm true}$ and position $p_{\rm true}$.
\item $PSF(p|p_{\rm true}, E_{\rm true})$ is the point spread function (PSF). It gives the probability of
  measuring a direction $p$ when the true direction is $p_{\rm true}$ and the true energy is $E_{\rm true}$.
  \gammaray instruments consider the probability density of the angular separation between true and reconstructed directions
  $\delta p = p_{\rm true} - p$, i.e. $PSF(\delta p|p_{\rm true}, E_{\rm true})$.
\item $E_{\rm disp}(E|p_{\rm true}, E_{\rm true})$ is the energy dispersion. It gives the probability to
  reconstruct the photon at energy $E$ when the true energy is $E_{\rm true}$ and the true position $p_{\rm true}$.
  \gammaray instruments consider the probability density of the migration $\mu=\frac{E}{E_{\rm true}}$,
  i.e. $E_{\rm disp}(\mu|p_{\rm true}, E_{\rm true})$.
\end{itemize}

@@

\subsection{\gammapy data analysis workflow}
The first step in \gammaray data analysis is the selection and extraction of observations
based of their metadata including information such as pointing direction, observation
time and observation conditions. The access to the events data and instrument 
reponse per observation is supported by classes and methods
in the \code{gammapy.data} (see Section~\ref{ssec:gammapy-data}) and the \code{gammapy.irf}
(see Section~\ref{ssec:gammapy-irf}) subpackages.

@@

%% CN: for 1D analysis datasets, DL4 can actually be written as OGIP files,
%% so as reduced datasets in a general format, not specific to Gammapy.
%% Maybe you can briefly mention that this is due to the fact that the 1D Poisson
%% spectral analysis is liong consilidated, but this also means that Gammapy 's
%% dataset can handle lower energy data, e.g. X-ray.
%% We also said in the abstract that Gammapy is compliant with other specifications
%% adopted in high-energy astrophysics.
The counts data and the reduced IRFs in the form of maps are bundled into datasets
that represent the fourth data level (DL4). These reduced datasets can be written to disk,
in a format specific to \gammapy to allow users to read them back at any time later
for modeling and fitting. Different variations of such datasets support different 
analysis methods and fit statistics. The datasets can be used to perform a joint-likelihood
fit, allowing one to combine different measurements, e.g. from different observations
but also from different instruments or event classes. They can also be used for binned
simulation as well as event sampling to simulate DL3 events data.
The various DL4 objects and the associated functionalities are
implemented in the \code{gammapy.datasets} subpackage (see Section~\ref{ssec:gammapy-datasets}).

The next step is then typically to model and fit the datasets, either
individually, or in a joint likelihood analysis. For this purpose \gammapy
provides a uniform interface to multiple fitting backends. In addition to
providing a variety of built-in models, including spectral,
spatial and temporal model classes to describe the \gammaray emission in the sky,
custom user-defined models are also supported.
Spectral models can be simple analytical models or more complex ones from radiation
mechanisms of accelerated particle populations (e.g. inverse Compton or $\pi^{0}$ decay).
Independently or subsequently to the global modelling, the data can be
re-grouped to compute flux points, light curves and flux as well as significance
maps in different energy bands.
The modelling and fitting functionalities are implemented in the \code{gammapy.modeling},
\code{gammapy.estimators} and \code{gammapy.stats} subpackages (see respectively
Section~\ref{ssec:gammapy-modeling}, \ref{ssec:gammapy-estimators} and \ref{ssec:gammapy-stats}).

@@

Figure~\ref{fig:data_flow} shows an overview of the different sub-packages and
their relation to each other. The \code{gammapy.data} and \code{gammapy.irf}
sub-packages define data objects to represent DL3 data, such as
event lists and IRFs as well as functionality
to read the DL3 data from disk into memory. The \code{gammapy.makers} sub-package
contains the functionality to reduce the DL3 data to binned maps.
Binned maps and datasets, which represent a collection of binned
maps, are defined in the \code{gammapy.maps} and \code{gammapy.datasets}
sub-packages, respectively. Parametric models, which are defined in
\code{gammapy.modeling}, are used to jointly model a combination
of datasets, for example, to compute a spectrum using data from several facilities. Estimator classes,
which are contained in \code{gammapy.estimators}, are used to
compute higher level science products such as flux and signficance maps,
light curves or flux points. Finally there is a \code{gammapy.analysis}
sub-package which provides a high-level interface for executing analyses
defined from configuration files. In the following sections we will
introduce all sub-packages and their functionalities in more detail.

\subsection{gammapy.data}
\label{ssec:gammapy-data}
The \code{gammapy.data} sub-package implements the functionality to select,
read, and represent DL3 \gammaray data in memory. It provides the main user
interface to access the lowest data level. \gammapy currently only
supports data that is compliant with \code{v0.2} and \code{v0.3} of the \gadf data format.
DL3 data are typically bundled into individual observations, which
corresponds to stable periods of data acquisition. For IACT data analysis,
for which the GADF data model and Gammapy were initially conceived,
these are usually $20 - 30\,{\rm min}$ long.
Each observation is assigned a unique integer ID for reference.

A typical usage example is shown in Figure~\ref{fig*:minted:gp_data}.
First a \code{DataStore} object is created from the path of the data
directory. The directory contains an observation as well as a FITS HDU 
index file which assigns the correct data and IRF FITS files and HDUs
to the given observation ID. The \code{DataStore}
object gathers a collection of observations and provides ancillary
files containing information about the telescope observation mode and the
content of the data unit of each file. The \code{DataStore} allows for
selecting a list of observations based on specific filters.

@@

\subsection{gammapy.irf}
\label{ssec:gammapy-irf}
%
%
\begin{figure*}[t]
	\centering
	\includegraphics[width=1.\textwidth]{figures/irfs.pdf}
	\caption{
		Using \code{gammapy.irf} to read and plot instrument response functions.
		The left panel shows the effective area as a function of energy for
		the \cta, \hess, \magic, \hawc and \fermi instruments. The right panel shows
		the $68\%$ containment radius of the PSF as a function of energy for the \cta, \hess
		and \fermi instruments. The \cta IRFs are from the \textit{prod5} production. The \hess IRFs are from the DL3 DR1,
        using observation ID 033787. The \magic effective area is computed for a
        $20\,{\rm min}$ observation at the Crab Nebula coordinates. The
		\fermi IRFs use \textit{pass8} data and are also taken at the position of the Crab Nebula.
		The \hawc effective area is shown for the event classes $N_{Hit}=5 - 9$ as light gray
		lines along with the sum of all event classes as a black line. The \hawc IRFs are taken from
		the first public release of event data the \hawc collaboration. All IRFs do not correspond
		to the latest performance of the instruments, but still are representative of the 
		detector type and energy range. We exclusively relied on publicly available
		data provided by the collaborations. The data is also availabe in the
		\code{gammapy-data} repository.
    }
	\label{fig:irfs}
\end{figure*}
%

@@

\subsubsection{Effective Area}
\gammapy provides the class \code{EffectiveAreaTable2D} to
manage the effective area, which is usually defined in terms of true energy and offset angle.
The class functionalities offer the possibility to read from files or to create
it from scratch. The \code{EffectiveAreaTable2D} class can also convert, interpolate,
write, and evaluate the effective area for a given energy and offset angles, or
even plot the multi-dimensional effective area table.

@@

To handle the change of the PSF with the observational offset during the analysis 
the \code{PSFMap} class is used. It stores the radial profile of the PSF
depending on the true energy and position on the sky. During the modeling
step in the analysis, the PSF profile for each model component is 
looked up at its current position and converted into a 3d convolution kernel
which is used for the prediction of counts from that model component.

\subsubsection{Energy Dispersion}
For \iacts, the energy resolution and bias, sometimes called energy dispersion,
is typically parametrised in terms of the so-called
migration parameter ($\mu$), which is defined as the ratio between the
reconstructed energy and the true energy. By definition, the mean of this ratio is
close to unity for a small energy bias and its distribution can 
be typically described by a Gaussian profile. However, more complex
shapes are also common. The migration parameter is given at each offset angle and
reconstructed energy. The main sub-classes are the \code{EnergyDispersion2D} which is
designed to handle the raw instrument description, and the \code{EDispKernelMap},
which contains an energy disperion matrix per sky position. I.e., a 4-dimensional
sky map where at each position is associated to an energy dispersion matrix.
The energy dispersion matrix is a representation of the energy resolution
as a function of the true energy only and implemnted in \gammapy
by the sub-class \code{EDispKernel}.

@@

	\caption{
        Using \code{gammapy.maps} to create a WCS, a HEALPix and a region
		based data structures. The initialisation parameters include
        consistently the positions of the center of the map, the pixel
        size, the extend of the map as well as the energy axis definition.
        The energy minimum and maximum values for the creation of the
        \code{MapAxis} object can be defined as strings also specifying the
        unit. Region definitions can be passed as strings following
        the DS9 region specifications \url{http://ds9.si.edu/doc/ref/region.html}.The output
		of the code example is shown in Figure~\ref{fig:code_example_gp_maps}.
        }
    \label{fig*:minted:gp_maps}
\end{figure}

@@

\subsubsection{HEALPix Maps}
This pixelization scheme ~\citep{Calabretta2002} provides a
subdivision of a sphere in which each pixel covers the same surface area as
every other pixel. As a consequence, however, pixel shapes are no longer
rectangular, or regular.
This pixelisation is typically used for all-sky data, such as data
from the \hawc or \fermi observatory. \gammapy natively supports
the multiscale definition of the HEALPix pixelisation and thus
allows for easy up and downsampling of the data. In addition to
the all-sky map, \gammapy also supports a local HEALPix
pixelisation where the size of the map is constrained to a given
radius.
For local neighbourhood operations, such as convolution \gammapy relies
on projecting the HEALPix data to a local tangential WCS grid.
This data structure is represented by the \code{HpxGeom} and \code{HpxNDMap}
classes.

@@

The \code{SkyModel} class represents the factorised model in Eq.~\ref{eq:source_model_dependency}
(the spatial and temporal components being optional).
A \code{SkyModel} object can represent the sum of several emission components:
either, for example, from multiple sources and from a diffuse emission, or from several spectral
components within the same source. To handle lists of multiple \code{SkyModel} objects, \gammapy
implements a \code{Models} class.

>>>>

\abstract
	{
		Traditionally, TeV-\gammaray astronomy has been conducted
		by experiments employing proprietary data and analysis software.
		However, the next generation of \gammaray instruments,
		such as the Cherenkov Telescope Array Observatory (CTAO), will be operated as open observatories.
		Alongside the data, they will also make associated software tools available to a wider community.
    This necessity prompted the development of open, high-level astronomy software customised for high-energy astrophysics.
	}
	{
		In this article, we present \gammapy, an open-source Python package for the analysis of astronomical \gammaray data,
		and illustrate the functionalities of its first long-term-support release, version 1.0.
		Built on the modern Python scientific ecosystem, \gammapy provides a uniform platform for reducing and
		modelling data from different \gammaray instruments for many analysis scenarios.
    \gammapy complies with several well-established data conventions in high-energy astrophysics, providing serialised data products that
		are interoperable with other software packages.
	}
	{
		Starting from event lists and instrument response functions,
		\gammapy provides the functionalities for reducing data binned in energy and sky coordinates.
		Several techniques for background estimation are implemented in the package to handle the residual hadronic background.
		After the data are binned, the flux and morphology of one or more \gammaray sources can be estimated
		using Poisson maximum likelihood fitting
		and assuming a variety of spectral, temporal, and spatial models.
		Estimation of flux points, likelihood profiles and light curves is also supported.
	}
	{
		After describing the structure of the package, we show the capabilities of \gammapy
		in multiple traditional and novel \gammaray analysis scenarios using public data, 
		such as spectral and spectro-morphological modelling and estimations of a spectral energy
		distribution and a light curve.
		Its flexibility and its power are displayed in a final multi-instrument example,
		where datasets from different instruments, at different stages of data reduction,
		are simultaneously fitted with an astrophysical flux model.
	}{}

@@

%% \gammaray astronomy
%% CN: I think we should use the expression Water Cherenkov Detectors and use the
%% abbreviation (WCD) all along the text, later we use "particle samplers"
The \gammaray range of the electromagnetic spectrum provides us insights into the
most energetic processes in the universe such as those accelerating particles in the surroundings of
black holes, and remnants of supernova explosions. As in other
branches of astronomy, \gammarays can be observed by
satellite as well as ground-based instruments.
Ground-based instruments use the Earth's atmosphere as a particle detector.
Very-high-energy (VHE) cosmic \gammarays interact in the atmosphere and
create large showers of secondary particles that can be observed from the ground.
Ground-based \gammaray astronomy relies on these extensive air showers to detect the
primary \gammaray photons and infer their incident direction and energy.
VHE \gammaray astronomy covers the energy range from few tens of \si{GeV} up to the \si{PeV}.
There are two main categories of ground-based instruments:

@@

\textit{Water Cherenkov Detectors (WCDs)} detect particles directly from the tail of the
shower when it reaches the ground. These instruments have a very
large FoV, and large duty-cycle, but a higher energy threshold and
lower signal-to-noise ratios compared to IACTs~\citep{2015CRPhy..16..610D}.

\begin{figure*}[t]
	\centering
	\includegraphics[width=\textwidth]{figures/big-picture.pdf}
	\caption{
		Core idea and relation of \gammapy to different \gammaray instruments
		and the gamma astro data format (GADF). The top left shows the
		group of current and future pointing instruments based on the 
		imaging atmospheric Cherenkov technique (IACT). This includes
		instruments such as the Cherenkov Telescope Array Observatory (CTAO),
		the High Energy Stereoscopic System (H.E.S.S.), the
		Major Atmospheric Gamma Imaging Cherenkov Telescopes (MAGIC),
		and the Very Energetic Radiation Imaging Telescope Array System (VERITAS).
		The lower left shows the group of all-sky instruments such as the
		Fermi Large Area Telescope (Fermi-LAT) and the High Altitude
		Water Cherenkov Observatory (HAWC). The calibrated data of all those
		instruments can be converted and stored into the common GADF data format.
		\gammapy can read data stored in the GADF format.
		The \gammapy package is a community-developed project that
		provides a common interface to the data and analysis of all
		these \gammaray instruments. This way users can also easily combine data from
		different instruments and perform a joint analysis.
		\gammapy is built on the scientific Python ecosystem, and the required dependencies
		are shown below the \gammapy logo.
	}
	\label{fig:big_picture}

@@

%TODO: better transition into context
The Cherenkov Telescope Array Observatory (CTAO) will be the first ground-based \gammaray instrument to be operated as an open observatory.
Its high-level data\footnote{The lowest reduction level of data published by CTAO will be reconstructed event lists and corresponding instrument response functions.} will be shared publicly after
some proprietary period, and the software required to analyze it will be distributed
as well. To allow the re-usability of data from existing instruments and their interoperability,
it is required to use open data formats and open tools that can support the various analysis methods
commonly used in the field.

@@

%% \gammapy: concept and goals
%%% Question: starting date? Is is relevant, what is the real start?
The \gammapy project was started following the model of \astropy, with the objective of building a common
software library for very high-energy \gammaray data analysis \citep{gammapy_2015}. 
The core of the idea is illustrated in Figure~\ref{fig:big_picture}. The various \gammaray instruments
can export their data to a common data format (\gadf) and then these data can be combined and
analysed using a common software library.
The \gammapy is package an independent community-developed software project, it has been selected to be the core library for the Science Analysis tools of CTAO
but also involves contributors associated to other instruments.
The \gammapy package is built on the scientific Python ecosystem: it uses \numpy~\citep{numpy} for n-dimensional data
structures, \scipy~\citep{2020SciPy-NMeth} for numerical algorithms, \astropy~\citep{astropy} for
astronomy-specific functionality, \iminuit~\citep{iminuit} for numerical minimisation
and \matplotlib~\citep{matplotlib} for visualization.

%% CN: what about mentioning the joint Crab? or is it mentioned later?
With the public availability of the \gadf format specifications and the
\gammapy package, some experiments started to make limited subsets of
their \gammaray data publicly available for testing and validating
\gammapy. For example, the \hess collaboration released a limited test
dataset (about 50 hours of observations taken between 2004 and 2008)
based on the \gadf DL3 format \citep{HESS_DR1}. This data release served
as a basis for validation of open analysis tools, including \gammapy 
\cite[see e.g.][]{Mohrmann2019}. The \hawc collaboration also released
a limited test dataset of the Crab Nebula, which was used to validate
the \gammapy package in \cite{Olivera2022}.
The increased availability of public data that followed the definition of a common data format, and the development of \gammapy as a community-driven open software, led the way toward a more open science in the very-high-energy \gray Astronomy domain. In future CTAO will be an open observatory committed to follow the FAIR (Findable, Accessible, Interoperable and Reusable) principles \citep{FAIR16, FAIR22} that define the key requirements for open science.

@@

\section{Gamma-ray Data Analysis}
\begin{figure*}[t]
	\centering
	\includegraphics[width=1.\textwidth]{figures/data_flow.pdf}
	\caption{
		\gammapy sub-package structure and data analysis workflow. The top row
        defines the different levels of data reduction, from lists of \gammaray-like
        events on the left (DL3), to high-level scientific products
        (DL5) on the right. The direction of the data flow is illustrated with the
        gray arrows. The gray folder icons represent the different sub-packages
        in \gammapy and their names. Below each icon there is a list of the most
        important objects defined in the sub-package.
    }
	\label{fig:data_flow}
\end{figure*}
%
\label{sec:gammaray-data-analysis}
% This might not be the bast place compared to introduction
The data analysis process in \gammaray astronomy is usually split into two parts.
The first one deals with the data processing from detector measurement, calibration, event
reconstruction and selection to yield a list of reconstructed \gammaray event candidates.
This part of the data reduction sequence, sometimes referred to as low-level analysis,
is usually very specific to a given observation technique and even to a given instrument.

The other sequence, referred to as high-level analysis, deals with the extraction of physical
quantities related to \gammaray sources and the production of high-level products such as spectra,
light curves and catalogs. The methods applied here are more generic and are broadly
shared across the field. The similarity in the high-level analysis would also allow
for combining data from multiple instruments.

@@

where:
\begin{itemize}
\setlength\itemsep{1em}
\item $A_{\rm eff}(p_{\rm true}, E_{\rm true})$ is the effective collection area of the detector. It is the product
  of the detector collection area times its detection efficiency at true energy $E_{\rm true}$ and position $p_{\rm true}$.
\item $PSF(p|p_{\rm true}, E_{\rm true})$ is the point spread function (PSF). It gives the probability of
  measuring a direction $p$ when the true direction is $p_{\rm true}$ and the true energy is $E_{\rm true}$.
  \gammaray instruments consider $PSF(\delta p|p_{\rm true}, E_{\rm true})$, the probability density of the angular separation between true and reconstructed directions,
  $\delta p = p_{\rm true} - p$.
\item $E_{\rm disp}(E|p_{\rm true}, E_{\rm true})$ is the energy dispersion. It gives the probability to
  reconstruct the photon at energy $E$ when the true energy is $E_{\rm true}$ and the true position $p_{\rm true}$.
  \gammaray instruments consider $E_{\rm disp}(\mu|p_{\rm true}, E_{\rm true})$, the probability density of the migration, $\mu=\frac{E}{E_{\rm true}}$.
\end{itemize}

@@

\subsection{\gammapy data analysis workflow}
The first step in \gammaray data analysis is the selection and extraction of observations
based on their metadata including information such as pointing direction, observation
time and observation conditions. The access to the events data and instrument 
reponse per observation is supported by classes and methods
in the \code{gammapy.data} (see Section~\ref{ssec:gammapy-data}) and the \code{gammapy.irf}
(see Section~\ref{ssec:gammapy-irf}) subpackages.

@@

%% CN: for 1D analysis datasets, DL4 can actually be written as OGIP files,
%% so as reduced datasets in a general format, not specific to Gammapy.
%% Maybe you can briefly mention that this is due to the fact that the 1D Poisson
%% spectral analysis is liong consilidated, but this also means that Gammapy 's
%% dataset can handle lower energy data, e.g. X-ray.
%% We also said in the abstract that Gammapy is compliant with other specifications
%% adopted in high-energy astrophysics.
The counts data and the reduced IRFs in the form of maps are bundled into datasets
that represent the fourth data level (DL4). These reduced datasets can be written to disk,
in a format specific to \gammapy to allow users to read them back at any time later
for modelling and fitting. Different variations of such datasets support different 
analysis methods and fit statistics. The datasets can be used to perform a joint-likelihood
fit, allowing one to combine different measurements, e.g. from different observations
but also from different instruments or event classes. They can also be used for binned
simulation as well as event sampling to simulate DL3 events data.
The various DL4 objects and the associated functionalities are
implemented in the \code{gammapy.datasets} subpackage (see Section~\ref{ssec:gammapy-datasets}).

The next step is then typically to model and fit the datasets, either
individually, or in a joint likelihood analysis. For this purpose \gammapy
provides a uniform interface to multiple fitting backends. In addition to
providing a variety of built-in models, including spectral,
spatial and temporal model classes to describe the \gammaray emission in the sky,
custom user-defined models are also supported.
Spectral models can be simple analytical models or more complex ones from radiation
mechanisms of accelerated particle populations (e.g. inverse Compton or $\pi^{0}$ decay).
Independently or subsequently to the global modelling, the data can be
re-grouped to compute flux points, light curves and flux maps as well as significance
maps in different energy bands.
The modelling and fitting functionalities are implemented in the \code{gammapy.modeling},
\code{gammapy.estimators} and \code{gammapy.stats} subpackages (see respectively
Section~\ref{ssec:gammapy-modeling}, \ref{ssec:gammapy-estimators} and \ref{ssec:gammapy-stats}).

@@

Figure~\ref{fig:data_flow} shows an overview of the different sub-packages and
their relation to each other. The \code{gammapy.data} and \code{gammapy.irf}
sub-packages define data objects to represent DL3 data, such as
event lists and IRFs as well as functionality
to read the DL3 data from disk into memory. The \code{gammapy.makers} sub-package
contains the functionality to reduce the DL3 data to binned maps.
Binned maps and datasets, which represent a collection of binned
maps, are defined in the \code{gammapy.maps} and \code{gammapy.datasets}
sub-packages, respectively. Parametric models, which are defined in
\code{gammapy.modeling}, are used to jointly model a combination
of datasets, for example, to compute a spectrum using data from several facilities. Estimator classes,
which are contained in \code{gammapy.estimators}, are used to
compute higher level science products such as flux and signficance maps,
light curves or flux points. Finally there is a \code{gammapy.analysis}
sub-package which provides a high-level interface for executing analyses
defined from configuration files. In the following sections, we will
introduce all sub-packages and their functionalities in more detail.

\subsection{gammapy.data}
\label{ssec:gammapy-data}
The \code{gammapy.data} sub-package implements the functionality to select,
read, and represent DL3 \gammaray data in memory. It provides the main user
interface to access the lowest data level. \gammapy currently only
supports data that is compliant with \code{v0.2} and \code{v0.3} of the \gadf data format.
DL3 data are typically bundled into individual observations, corresponding to stable periods of data acquisition. For IACT data analysis,
for which the GADF data model and Gammapy were initially conceived,
these are usually $20 - 30\,{\rm min}$ long.
Each observation is assigned a unique integer ID for reference.

A typical usage example is shown in Figure~\ref{fig*:minted:gp_data}.
First a \code{DataStore} object is created from the path of the data
directory. The directory contains an observation as well as a FITS HDU \footnote{Header Data Unit} 
index file which assigns the correct data and IRF FITS files and HDUs
to the given observation ID. The \code{DataStore}
object gathers a collection of observations and provides ancillary
files containing information about the telescope observation mode and the
content of the data unit of each file. The \code{DataStore} allows for
selecting a list of observations based on specific filters.

@@

\subsection{gammapy.irf}
\label{ssec:gammapy-irf}
%
%
\begin{figure*}[t]
	\centering
	\includegraphics[width=1.\textwidth]{figures/irfs.pdf}
	\caption{
		Using \code{gammapy.irf} to read and plot instrument response functions.
		The left panel shows the effective area as a function of energy for
		the \cta, \hess, \magic, \hawc and \fermi instruments. The right panel shows
		the $68\%$ containment radius of the PSF as a function of energy for the \cta, \hess
		and \fermi instruments. The \cta IRFs are from the \textit{prod5} production. The \hess IRFs are from the DL3 DR1,
        using observation ID 033787. The \magic effective area is computed for a
        $20\,{\rm min}$ observation at the Crab Nebula coordinates. The
		\fermi IRFs use \textit{pass8} data and are also taken at the position of the Crab Nebula.
		The \hawc effective area is shown for the event classes $N_{Hit}=5 - 9$ as light gray
		lines along with the sum of all event classes as a black line. The \hawc IRFs are taken from
		the first public release of events data by the \hawc collaboration. All IRFs do not correspond
		to the latest performance of the instruments, but still are representative of the 
		detector type and energy range. We exclusively relied on publicly available
		data provided by the collaborations. The data is also available in the
		\code{gammapy-data} repository.
    }
	\label{fig:irfs}
\end{figure*}
%

@@

\subsubsection{Effective Area}
\gammapy provides the class \code{EffectiveAreaTable2D} to
manage the effective area, which is usually defined in terms of true energy and offset angle.
The class functionalities offer the possibility to read from files or to create
it from scratch. The \code{EffectiveAreaTable2D} class can also convert, interpolate,
write, and evaluate the effective area for a given energy and offset angle, or
even plot the multi-dimensional effective area table.

@@

To handle the change of the PSF with the observational offset during the analysis 
the \code{PSFMap} class is used. It stores the radial profile of the PSF
depending on the true energy and position on the sky. During the modelling
step in the analysis, the PSF profile for each model component is 
looked up at its current position and converted into a 3d convolution kernel
which is used for the prediction of counts from that model component.

\subsubsection{Energy Dispersion}
For \iacts, the energy resolution and bias, sometimes called energy dispersion,
is typically parametrised in terms of the so-called
migration parameter ($\mu$), which is defined as the ratio between the
reconstructed energy and the true energy. By definition, the mean of this ratio is
close to unity for a small energy bias and its distribution can 
be typically described by a Gaussian profile. However, more complex
shapes are also common. The migration parameter is given at each offset angle and
reconstructed energy. The main sub-classes are the \code{EnergyDispersion2D} which is
designed to handle the raw instrument description, and the \code{EDispKernelMap},
which contains an energy disperion matrix per sky position. I.e., a 4-dimensional
sky map where each position is associated to an energy dispersion matrix.
The energy dispersion matrix is a representation of the energy resolution
as a function of the true energy only and implemented in \gammapy
by the sub-class \code{EDispKernel}.

@@

	\caption{
        Using \code{gammapy.maps} to create a WCS, a HEALPix and a region
		based data structures. The initialisation parameters include
        consistently the positions of the center of the map, the pixel
        size, the extend of the map as well as the energy axis definition.
        The energy minimum and maximum values for the creation of the
        \code{MapAxis} object can be defined as strings also specifying the
        unit. Region definitions can be passed as strings following
        the DS9 region specifications \url{http://ds9.si.edu/doc/ref/region.html}. The output
		of the code example is shown in Figure~\ref{fig:code_example_gp_maps}.
        }
    \label{fig*:minted:gp_maps}
\end{figure}

@@

\subsubsection{HEALPix Maps}
This pixelization scheme ~\citep{Calabretta2002} provides a
subdivision of a sphere in which each pixel covers the same surface area as
every other pixel. As a consequence, however, pixel shapes are no longer
rectangular, or regular.
This pixelisation is typically used for all-sky data, such as data
from the \hawc or \fermi observatory. \gammapy natively supports
the multiscale definition of the HEALPix pixelisation and thus
allows for easy upsampling and downsampling of the data. In addition to
the all-sky map, \gammapy also supports a local HEALPix
pixelisation where the size of the map is constrained to a given
radius.
For local neighbourhood operations, such as convolution, \gammapy relies
on projecting the HEALPix data to a local tangential WCS grid.
This data structure is represented by the \code{HpxGeom} and \code{HpxNDMap}
classes.

@@

The \code{SkyModel} class represents the factorised model in Eq.~\ref{eq:source_model_dependency}
(the spatial and temporal components being optional).
A \code{SkyModel} object can represent the sum of several emission components:
either, for example, from multiple sources and from a diffuse emission, or from several spectral
components within the same source. To handle a list of multiple \code{SkyModel} objects, \gammapy
implements a \code{Models} class.
